---
title: "HumanPilot: Clustering using PCA, UMAP, marker genes, spatial coordinates"
author: "Lukas Weber"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

This script contains code for several attempts at clustering the HumanPilot brain dataset, using various combinations of top PCs, UMAP components, marker genes, and spatial coordinates.

The reasoning is that performing clustering on a combined set of molecular feature dimensions and spatial dimensions (e.g. concatenating the two spatial dimensions as additional columns with the top PCs) is a simple and intuitive way to combine the molecular and spatial data, which should improve the ability of the clustering algorithms to follow the spatial layer structure.

However, it is not clear how best to balance the molecular dimensions (e.g. top PCs, top UMAP coordinates, top marker genes) with the two spatial dimensions. Questions include:

- What scaling should be used for the spatial dimensions? The units should be approximately comparable, so one possibility could be to scale the two spatial dimensions to approximately the same range as the top PC or the top UMAP coordinate. Also, z-scaling should not be used: for the molecular dimensions this would scale up less meaningful dimensions, while for spatial dimensions z-scaling does not make sense since these exist on a uniform scale with physical meaning.

- How many molecular dimensions should be used? E.g. top 50 PCs, top 10 UMAPs, top 10 marker genes, etc. In particular there seems to be a risk that if too many molecular dimensions are used (e.g. top 50 PCs or all 1942 highly variable genes), then the information contained in the spatial dimensions will be swamped by the molecular dimensions.

- For the UMAP clustering: here the argument is that using the top few UMAP coordinates is a simple way to reduce as much as possible of the molecular information into a smaller number of dimensions, to avoid swamping the spatial information by including a larger number of PCs or HVGs. Then e.g. the top 5-10 UMAP dimensions can be combined with the two spatial dimensions on a more equal basis.


# Code

The following code runs clustering using various combinations of top PCs, UMAP components, marker genes, and spatial coordinates. The main parameters and other design choices are:

- whether to use PCs, UMAP coordinates, or marker genes for the molecular information
- how many PCs, UMAP coordinates, or marker genes
- how to scale the spatial coordinates (e.g. similar range as the top UMAP coordinate)
- number of clusters
- type of clustering algorithm (e.g. graph-based makes more sense than k-means since the clusters are not ellipsoidal in the two spatial dimensions)


```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
suppressPackageStartupMessages(library(uwot))
suppressPackageStartupMessages(library(scran))
suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(RColorBrewer))
```


## Load data

Load SingleCellExperiment data object containing UMI counts, top 50 PCs, and spatial coordinates.

Note: using data from one sample only for now (sample 151673).

```{r}
# load scran output file
load("../../data/Human_DLPFC_Visium_processedData_sce_scran.Rdata")
sce
```

```{r}
# select spots from one sample
ix_151673 <- colData(sce)$sample_name == 151673

sce_151673 <- sce[, ix_151673]
sce_151673
```


## Extract features

Extract features from object: top 50 PCs, spatial coordinates, log counts of marker genes detected using SpatialDE (including some for white matter).

```{r}
# extract top 50 PCs
dims_pcs <- reducedDim(sce_151673, type = "PCA")
stopifnot(nrow(dims_pcs) == ncol(sce_151673))


# extract spatial dimensions
# note: y coordinate is reversed
xy_coords <- data.frame(
    x_coord = colData(sce_151673)[, c("imagecol")], 
    y_coord = -colData(sce_151673)[, c("imagerow")]
)
dims_spatial <- xy_coords
stopifnot(nrow(dims_spatial) == ncol(sce_151673))


# extract log counts of marker genes from SpatialDE
marker_genes_SpatialDE <- c(
    "NEFH", "PCP4", "HPCAL1", "VSNL1", "TMSB10", "CCK", "HOPX", "PPP3CA", "GPM6A", 
    "ENC1", "NEFM", "NEFL", "DIRAS2", "SNCG", "FBXL16", "GFAP", "CAMK2N1", "CXCL14", 
    "TUBB2A", "OLFM1", "MALAT1", "APLP1", 
    "MOBP", "ERMN", "CLDND1", "TF"
)
length(marker_genes_SpatialDE)

# identify using gene symbols
ix <- match(marker_genes_SpatialDE, rowData(sce_151673)$gene_name)
length(ix)
marker_genes_SpatialDE_sym <- rowData(sce_151673)$gene_id[ix]
marker_genes_SpatialDE_sym

dims_markers_SpatialDE <- t(as.matrix(logcounts(sce_151673)[marker_genes_SpatialDE_sym, , drop = FALSE]))

dims_markers_SpatialDE[1:6, 1:6]
dim(dims_markers_SpatialDE)

stopifnot(nrow(dims_markers_SpatialDE) == ncol(sce_151673))
```


## Run UMAP

Run UMAP: the aim will be to use the top 5-10 UMAP components for clustering.

Note: running UMAP on the top 50 PCs for faster runtime. To do: run on top 1942 highly variable genes instead.

Note: running UMAP on one sample only; could also run on all samples combined.

```{r}
# keep top 50 UMAP components
set.seed(123)
dims_umap <- umap(dims_pcs, n_components = 50)

stopifnot(nrow(dims_umap) == ncol(sce_151673))
```


## Run UMAP on marker genes

Another alternative: run UMAP on the set of SpatialDE marker genes.

```{r}
# keep top 10 UMAP components
set.seed(123)
dims_umap_markers_SpatialDE <- umap(dims_markers_SpatialDE, n_components = 10)

stopifnot(nrow(dims_umap_markers_SpatialDE) == ncol(sce_151673))
```


## Scale spatial dimensions

Need all dimensions to be on approximately comparable scales.

A simple way to achieve this is to scale the spatial dimensions to a range comparable to the range of the top UMAPs or PCs. However the results will be highly sensitive to the exact choice of range.

Note: do not use z-scaling for UMAP, PCs, or spatial dimensions (this would scale up less meaningful dimensions for UMAP or PCs; and doesn't make sense for physical spatial coordinates). However, marker gene log counts need to be z-scaled.

```{r}
# check range of UMAP dimensions
summary(dims_umap[, 1:5])

mean(dims_umap[, 1])
sd(dims_umap[, 1])
max(abs(dims_umap))

range(dims_umap[, 1])
range(dims_umap[, 2])
range(dims_umap[, 3])

colnames(dims_umap) <- paste0("UMAP_", seq_len(ncol(dims_umap)))
```


```{r}
# check range of PCs
summary(dims_pcs[, 1:5])

mean(dims_pcs[, 1])
sd(dims_pcs[, 1])
max(abs(dims_pcs))

range(dims_pcs[, 1])
range(dims_pcs[, 2])
range(dims_pcs[, 3])
```


```{r}
# z-scaling for log counts of marker genes
# note: these are not ranked in any order
summary(dims_markers_SpatialDE[, 1:5])

dims_markers_SpatialDE <- scale(dims_markers_SpatialDE)

summary(dims_markers_SpatialDE[, 1:5])

mean(dims_markers_SpatialDE[, 1])
sd(dims_markers_SpatialDE[, 1])

max(abs(dims_markers_SpatialDE))
```


```{r}
# scale spatial dimensions
summary(as.data.frame(dims_spatial))
range(dims_spatial[, 1])
range(dims_spatial[, 2])

dims_spatial <- apply(as.matrix(dims_spatial), 2, function(col) {
    (col - min(col)) / (max(col) - min(col)) * 20 - 10
})

colnames(dims_spatial) <- c("spatial_x", "spatial_y")

summary(dims_spatial)

stopifnot(nrow(dims_spatial) == ncol(sce_151673))
```


```{r}
# alternative ranges for additional plots
dims_spatial2 <- apply(as.matrix(dims_spatial), 2, function(col) {
    (col - min(col)) / (max(col) - min(col)) * 10 - 5
})
summary(dims_spatial2)

dims_spatial3 <- apply(as.matrix(dims_spatial), 2, function(col) {
    (col - min(col)) / (max(col) - min(col)) * 4 - 2
})
summary(dims_spatial3)

dims_spatial4 <- apply(as.matrix(dims_spatial), 2, function(col) {
    (col - min(col)) / (max(col) - min(col)) * 2 - 1
})
summary(dims_spatial4)
```


## Graph-based clustering

Run standard Bioconductor graph-based clustering on subset of molecular dimensions and the two scaled spatial dimensions.

Note: graph-based clustering seems better suited than k-means for this dataset, since the layers do not have an ellipsoidal shape in the spatial dimensions. However, for other datasets, e.g. cancer data, k-means could also work.

Each clustering and plot below uses a different combination of molecular feature dimensions and spatial dimensions.

To do: arrange more systematically; e.g. to show impact of choice of range for scaling spatial dimensions (e.g. larger range tends to give more circular clusters, since spatial information dominates molecular information).

To do: also show impact of number of clusters (do clusters split up into more layers, or into more circular type clusters? from preliminary plots it seemed they tend to split into circular clusters first).


### Top PCs plus spatial dimensions

**Top 10 PCs plus 2 spatial dimensions, with spatial dimensions scaled to range +10 to -10.**

```{r, fig.width=6, fig.height=6}
# number of PCs
n_pcs <- 10
dims_clus <- cbind(dims_pcs[, seq_len(n_pcs), drop = FALSE], dims_spatial)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on top few PCs plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_PCs_spatial.png"
ggsave(filename, width = 6, height = 6)
```


**Top 10 PCs plus 2 spatial dimensions, with spatial dimensions scaled to range +5 to -5.**

```{r, fig.width=6, fig.height=6}
# number of PCs
n_pcs <- 10
dims_clus <- cbind(dims_pcs[, seq_len(n_pcs), drop = FALSE], dims_spatial2)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on top few PCs plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_PCs_spatial2.png"
ggsave(filename, width = 6, height = 6)
```


**Top 10 PCs without any spatial dimensions.**

```{r, fig.width=6, fig.height=6}
# number of PCs
n_pcs <- 50
dims_clus <- cbind(dims_pcs[, seq_len(n_pcs), drop = FALSE])
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on top few PCs")

filename <- "../plots/clustering/clustering_PCs.png"
ggsave(filename, width = 6, height = 6)
```


### Top UMAPs plus spatial dimensions

**Top 10 UMAPs plus 2 spatial dimensions, with spatial dimensions scaled to range +10 to -10.**

```{r, fig.width=6, fig.height=6}
# number of UMAPs
n_umap <- 10
dims_clus <- cbind(dims_umap[, seq_len(n_umap), drop = FALSE], dims_spatial)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on top few UMAPs plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_UMAPs_spatial.png"
ggsave(filename, width = 6, height = 6)
```


**Top 10 UMAPs plus 2 spatial dimensions, with spatial dimensions scaled to range +5 to -5.**

```{r, fig.width=6, fig.height=6}
# number of UMAPs
n_umap <- 10
dims_clus <- cbind(dims_umap[, seq_len(n_umap), drop = FALSE], dims_spatial2)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on top few UMAPs plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_UMAPs_spatial2.png"
ggsave(filename, width = 6, height = 6)
```


**Top 10 UMAPs without any spatial dimensions.**

```{r, fig.width=6, fig.height=6}
# number of UMAPs
n_umap <- 10
dims_clus <- cbind(dims_umap[, seq_len(n_umap), drop = FALSE])
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on top few UMAPs")

filename <- "../plots/clustering/clustering_UMAPs_spatial3.png"
ggsave(filename, width = 6, height = 6)
```


### SpatialDE marker genes plus spatial dimensions

**SpatialDE marker genes plus 2 spatial dimensions, with spatial dimensions scaled to range +10 to -10.**

```{r, fig.width=6, fig.height=6}
# number of marker genes
ncol(dims_markers_SpatialDE)
n_markers <- ncol(dims_markers_SpatialDE)
dims_clus <- cbind(dims_markers_SpatialDE[, seq_len(n_markers), drop = FALSE], dims_spatial)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on SpatialDE markers plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_SpatialDE_spatial.png"
ggsave(filename, width = 6, height = 6)
```


**SpatialDE marker genes plus 2 spatial dimensions, with spatial dimensions scaled to range +5 to -5.**

```{r, fig.width=6, fig.height=6}
# number of marker genes
ncol(dims_markers_SpatialDE)
n_markers <- ncol(dims_markers_SpatialDE)
dims_clus <- cbind(dims_markers_SpatialDE[, seq_len(n_markers), drop = FALSE], dims_spatial2)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on SpatialDE markers plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_SpatialDE_spatial2.png"
ggsave(filename, width = 6, height = 6)
```


**SpatialDE marker genes plus 2 spatial dimensions, with spatial dimensions scaled to range +2 to -2.**

```{r, fig.width=6, fig.height=6}
# number of marker genes
ncol(dims_markers_SpatialDE)
n_markers <- ncol(dims_markers_SpatialDE)
dims_clus <- cbind(dims_markers_SpatialDE[, seq_len(n_markers), drop = FALSE], dims_spatial3)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on SpatialDE markers plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_SpatialDE_spatial3.png"
ggsave(filename, width = 6, height = 6)
```


**SpatialDE marker genes plus 2 spatial dimensions, with spatial dimensions scaled to range +1 to -1.**

```{r, fig.width=6, fig.height=6}
# number of marker genes
ncol(dims_markers_SpatialDE)
n_markers <- ncol(dims_markers_SpatialDE)
dims_clus <- cbind(dims_markers_SpatialDE[, seq_len(n_markers), drop = FALSE], dims_spatial4)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on SpatialDE markers plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_SpatialDE_spatial4.png"
ggsave(filename, width = 6, height = 6)
```


**SpatialDE marker genes without any spatial dimensions.**

```{r, fig.width=6, fig.height=6}
# number of marker genes
ncol(dims_markers_SpatialDE)
n_markers <- ncol(dims_markers_SpatialDE)
dims_clus <- cbind(dims_markers_SpatialDE[, seq_len(n_markers), drop = FALSE])
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on SpatialDE markers")

filename <- "../plots/clustering/clustering_SpatialDE.png"
ggsave(filename, width = 6, height = 6)
```


### UMAP performed on SpatialDE marker genes plus spatial dimensions

**UMAP performed on SpatialDE marker genes, plus 2 spatial dimensions, with spatial dimensions scaled to range +5 to -5.**

```{r, fig.width=6, fig.height=6}
# number of marker genes
ncol(dims_umap_markers_SpatialDE)
dims_clus <- cbind(dims_umap_markers_SpatialDE, dims_spatial2)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on UMAP on SpatialDE markers plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_UMAP_SpatialDE_spatial2.png"
ggsave(filename, width = 6, height = 6)
```


**UMAP performed on SpatialDE marker genes, plus 2 spatial dimensions, with spatial dimensions scaled to range +2 to -2.**

```{r, fig.width=6, fig.height=6}
# number of marker genes
ncol(dims_umap_markers_SpatialDE)
dims_clus <- cbind(dims_umap_markers_SpatialDE, dims_spatial3)
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on UMAP on SpatialDE markers plus 2 spatial dims (scaled)")

filename <- "../plots/clustering/clustering_UMAP_SpatialDE_spatial3.png"
ggsave(filename, width = 6, height = 6)
```


**UMAP performed on SpatialDE marker genes without any spatial dimensions.**

```{r, fig.width=6, fig.height=6}
# number of marker genes
ncol(dims_umap_markers_SpatialDE)
dims_clus <- dims_umap_markers_SpatialDE
head(dims_clus, 2)

# clustering: see OSCA book (note transpose; number of clusters)
g <- buildSNNGraph(t(dims_clus), k = 10, d = ncol(dims_clus))
g_walk <- igraph::cluster_walktrap(g)

# select number of clusters
n_clus <- 8
clus <- igraph::cut_at(g_walk, n = n_clus)
table(clus)
stopifnot(length(clus) == ncol(sce_151673))

# plot
d_plot <- cbind(xy_coords, cluster = as.factor(clus))

ggplot(d_plot, aes(x = x_coord, y = y_coord, color = cluster)) + 
    geom_point(size = 2, alpha = 0.5) + 
    coord_fixed() + 
    scale_color_brewer(palette = "Paired") + 
    theme_bw() + 
    ggtitle("Clustering on UMAP on SpatialDE markers")

filename <- "../plots/clustering/clustering_UMAP_SpatialDE.png"
ggsave(filename, width = 6, height = 6)
```


