---
title: "HumanPilot: Clustering using PCA, UMAP, marker genes, spatial coordinates"
author: "Lukas Weber"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
    html_document:
        toc: true
        toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

This script contains code for several attempts at clustering the HumanPilot brain dataset, using various combinations of top PCs, UMAP dimensions, SpatialDE marker genes, and spatial coordinates.


```{r}
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(uwot)
    library(scran)
    library(scater)
    library(ggplot2)
    library(RColorBrewer)
})
```


## Load data

Load SingleCellExperiment data object containing UMI counts, top 50 PCs, and spatial coordinates.

```{r}
# load scran output file
load("../../data/Human_DLPFC_Visium_processedData_sce_scran.Rdata")
sce

# get sample names
sample_names <- as.factor(unique(colData(sce)$sample_name))
```


## SpatialDE marker genes

List of SpatialDE marker genes to use for clustering (from separate script "Comparison_SpatialDE_genes_pooled.Rmd").

```{r}
marker_genes_SpatialDE <- c(
    "ARPP19", "CAMK2N1", "CARTPT", "CCK", "CNP", "CRYAB", "CXCL14", "DIRAS2", 
    "EEF1A1", "ENC1", "FAU", "FTL", "GFAP", "HOPX", "HPCAL1", "MALAT1", "MOBP", 
    "NEFH", "NEFM", "PCP4", "RPL12", "SERPINE2", "TF"
)
length(marker_genes_SpatialDE)
```


## Extract features, run UMAP, run clustering for each sample

Run separately for each sample:

- extract features from data object: top 50 PCs, spatial coordinates, log counts of SpatialDE marker genes

- run UMAP (on either top 50 PCs or SpatialDE marker genes) (note: using log counts; no additional z-scaling required)

- scale spatial dimensions (note: use range comparable to range of UMAP dimension; no z-scaling since these are physical coordinates)

- run graph-based clustering (on various combinations of PCs, UMAP dimensions, SpatialDE marker genes, and spatial dimensions) (Bioconductor graph-based clustering; note graph-based clustering seems better suited than k-means since clusters do not necessarily have an ellipsoidal shape in the spatial dimensions)

```{r}
# parameters
n_umap <- 10
max_spatial <- 1
n_clus_k <- 10
n_clus <- 8


d_plot <- data.frame()


# run separately for each sample
for (i in seq_along(sample_names)) {
    
    # -----------------------------
    # extract features and run UMAP
    # -----------------------------
    
    # select spots from this sample
    ix <- colData(sce)$sample_name == sample_names[i]
    table(ix)
    sce_sub <- sce[, ix]
    
    # extract x-y coordinates of spots (note: y coordinate is reversed)
    xy_coords <- data.frame(
        x_coord = colData(sce_sub)[, c("imagecol")], 
        y_coord = -colData(sce_sub)[, c("imagerow")]
    )
    dims_spatial <- xy_coords
    colnames(dims_spatial) <- c("spatial_x", "spatial_y")
    stopifnot(nrow(dims_spatial) == ncol(sce_sub))
    
    # select SpatialDE marker genes
    stopifnot(all(marker_genes_SpatialDE %in% toupper(rowData(sce_sub)$gene_name)))
    sce_sub <- sce_sub[toupper(rowData(sce_sub)$gene_name) %in% marker_genes_SpatialDE, ]
    dim(sce_sub)
    
    # extract top 50 PCs
    dims_pcs <- reducedDim(sce_sub, type = "PCA")
    stopifnot(nrow(dims_pcs) == ncol(sce_sub))
    dim(dims_pcs)
    
    # run UMAP on top 50 PCs
    # note: no z-scaling
    set.seed(123)
    dims_umap <- umap(dims_pcs, n_components = n_umap)
    colnames(dims_umap) <- paste0("UMAP_", seq_len(ncol(dims_umap)))
    
    # extract log counts for SpatialDE marker genes
    dims_markers_SpatialDE <- t(as.matrix(logcounts(sce_sub)))
    stopifnot(ncol(dims_markers_SpatialDE) == length(marker_genes_SpatialDE))
    colnames(dims_markers_SpatialDE) <- marker_genes_SpatialDE
    dim(dims_markers_SpatialDE)
    
    # run UMAP on SpatialDE marker genes
    # note: no z-scaling
    set.seed(123)
    dims_umap_SpatialDE <- umap(dims_markers_SpatialDE, n_components = n_umap)
    colnames(dims_umap_SpatialDE) <- paste0("UMAP_", seq_len(ncol(dims_umap)))
    
    # scale spatial dimensions
    apply(dims_spatial, 2, range)
    dims_spatial <- apply(as.matrix(dims_spatial), 2, function(col) {
        (col - min(col)) / (max(col) - min(col)) * (2 * max_spatial) - max_spatial
    })
    apply(dims_spatial, 2, range)
    stopifnot(nrow(dims_spatial) == ncol(sce_sub))
    
    
    # ----------
    # clustering
    # ----------
    
    # graph-based clustering (see Bioconductor OSCA book)
    
    
    # (i) clustering on top 50 PCs
    
    dims_clus <- dims_pcs
    head(dims_clus, 2)
    
    g <- buildSNNGraph(t(dims_clus), k = n_clus_k, d = ncol(dims_clus))
    g_walk <- igraph::cluster_walktrap(g)
    clus <- igraph::cut_at(g_walk, n = n_clus)
    table(clus)
    
    stopifnot(length(clus) == nrow(xy_coords))
    
    res <- cbind(
        xy_coords, 
        cluster = clus, 
        method = "top50_PCs", 
        sample_name = paste0("sample_", sample_names[i])
    )
    
    d_plot <- rbind(d_plot, res)
    
    
    # (ii) clustering on top 10 UMAP dimensions
    
    dims_clus <- dims_umap
    head(dims_clus, 2)
    
    g <- buildSNNGraph(t(dims_clus), k = n_clus_k, d = ncol(dims_clus))
    g_walk <- igraph::cluster_walktrap(g)
    clus <- igraph::cut_at(g_walk, n = n_clus)
    table(clus)
    
    stopifnot(length(clus) == nrow(xy_coords))
    
    res <- cbind(
        xy_coords, 
        cluster = clus, 
        method = "top10_UMAP", 
        sample_name = paste0("sample_", sample_names[i])
    )
    
    d_plot <- rbind(d_plot, res)
    
    
    # (iii) clustering on logcounts of SpatialDE marker genes
    
    dims_clus <- dims_markers_SpatialDE
    head(dims_clus, 2)
    
    g <- buildSNNGraph(t(dims_clus), k = n_clus_k, d = ncol(dims_clus))
    g_walk <- igraph::cluster_walktrap(g)
    clus <- igraph::cut_at(g_walk, n = n_clus)
    table(clus)
    
    stopifnot(length(clus) == nrow(xy_coords))
    
    res <- cbind(
        xy_coords, 
        cluster = clus, 
        method = "SpatialDE", 
        sample_name = paste0("sample_", sample_names[i])
    )
    
    d_plot <- rbind(d_plot, res)
    
    
    # (iv) clustering on top 10 UMAP dimensions plus scaled spatial dimensions
    
    dims_clus <- cbind(dims_umap, dims_spatial)
    head(dims_clus, 2)
    
    g <- buildSNNGraph(t(dims_clus), k = n_clus_k, d = ncol(dims_clus))
    g_walk <- igraph::cluster_walktrap(g)
    clus <- igraph::cut_at(g_walk, n = n_clus)
    table(clus)
    
    stopifnot(length(clus) == nrow(xy_coords))
    
    res <- cbind(
        xy_coords, 
        cluster = clus, 
        method = "top10_UMAP_spatial", 
        sample_name = paste0("sample_", sample_names[i])
    )
    
    d_plot <- rbind(d_plot, res)
    
    
    # (v) clustering on logcounts of SpatialDE marker genes plus scaled spatial dimensions
    
    dims_clus <- cbind(dims_markers_SpatialDE, dims_spatial)
    head(dims_clus, 2)
    
    g <- buildSNNGraph(t(dims_clus), k = n_clus_k, d = ncol(dims_clus))
    g_walk <- igraph::cluster_walktrap(g)
    clus <- igraph::cut_at(g_walk, n = n_clus)
    table(clus)
    
    stopifnot(length(clus) == nrow(xy_coords))
    
    res <- cbind(
        xy_coords, 
        cluster = clus, 
        method = "SpatialDE_spatial", 
        sample_name = paste0("sample_", sample_names[i])
    )
    
    d_plot <- rbind(d_plot, res)
}

d_plot$cluster <- as.factor(d_plot$cluster)
d_plot$sample_name <- as.factor(d_plot$sample_name)
```


## Generate plots

Separate panel of plots for each sample.

```{r, fig.width=12, fig.height=3.25}
for (i in seq_along(sample_names)) {
    
    d_plot_sub <- d_plot[d_plot$sample_name == paste0("sample_", sample_names[i]), ]
    
    print(
        ggplot(d_plot_sub, aes(x = x_coord, y = y_coord, color = cluster)) + 
            facet_wrap(~ method, nrow = 1) + 
            geom_point(size = 0.5, alpha = 0.5) + 
            coord_fixed() + 
            scale_color_brewer(palette = "Paired") + 
            theme_bw() + 
            ggtitle(paste0("Clustering: sample ", sample_names[i]))
    )
    
    filename <- paste0("../plots/clustering/clustering_sample_", sample_names[i], ".png")
    ggsave(filename, width = 12, height = 3.25)
}
```


