---
title: "SpatialDE subsampling comparison"
author: "Lukas Weber"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
    html_document:
        toc: true
        toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# SpatialDE subsampling comparison

Comparison of gene lists generated from Stephanie's SpatialDE runs with and without subsampling on the number of spots. (The version with subsampling used subsampling to 1500 spots per sample for faster runtime.)


## Load data

Load two sets of SpatialDE results for each sample:

- all spots (no subsampling)
- subsampling to 1500 spots per sample


```{r, message = FALSE}
library(SingleCellExperiment)
library(readr)
library(dplyr)
library(UpSetR)
```


```{r, message = FALSE, warning = FALSE}
# load spreadsheets containing SpatialDE results
spatialDE_files_all <- list.files(
    "../../data/SpatialDE/spatialDE_all", pattern = "\\.csv$", full.names = TRUE
)
spatialDE_files_sub <- list.files(
    "../../data/SpatialDE/spatialDE_subsampled", pattern = "\\.csv$", full.names = TRUE
)

spatialDE_results_all <- list()
spatialDE_results_sub <- list()

for (i in seq_along(spatialDE_files_all)) {
    spatialDE_results_all[[i]] <- read_csv(spatialDE_files_all[i])
}
for (i in seq_along(spatialDE_files_sub)) {
    spatialDE_results_sub[[i]] <- read_csv(spatialDE_files_sub[i])
}

sample_names_all <- gsub("_spatialDE.*$", "", gsub("^.*/", "", spatialDE_files_all))
sample_names_sub <- gsub("_spatialDE.*$", "", gsub("^.*/", "", spatialDE_files_sub))

stopifnot(all(sample_names_all == sample_names_sub))

sample_names <- sample_names_all

names(spatialDE_results_all) <- sample_names
names(spatialDE_results_sub) <- sample_names
```


## Comparison of SpatialDE gene lists

Compare the following:

- total number of genes per sample
- number of significant genes per sample (q-values < 0.05)
- number of genes per sample with q-values = 0

```{r}
# check numbers of genes

# total number of genes per sample (note numbers differ slightly)
sapply(spatialDE_results_all, nrow)
sapply(spatialDE_results_sub, nrow)

# number of significant genes per sample (q-values < 0.05)
sapply(spatialDE_results_all, function(d) table(d$qval < 0.05))
sapply(spatialDE_results_sub, function(d) table(d$qval < 0.05))

# number of genes per sample with q-values = 0
sapply(spatialDE_results_all, function(d) table(d$qval == 0))
sapply(spatialDE_results_sub, function(d) table(d$qval == 0))
```


```{r}
# extract gene lists

# significant genes per sample (q-values < 0.05)
spatialDE_sig_all <- lapply(spatialDE_results_all, function(d) {
    filter(d, qval < 0.05)$g
})
spatialDE_sig_sub <- lapply(spatialDE_results_sub, function(d) {
    filter(d, qval < 0.05)$g
})

# genes per sample with q-values = 0
spatialDE_zero_all <- lapply(spatialDE_results_all, function(d) {
    filter(d, qval == 0)$g
})
spatialDE_zero_sub <- lapply(spatialDE_results_sub, function(d) {
    filter(d, qval == 0)$g
})
```


```{r}
# convert gene symbols to gene names

# load scran output file from Leo
load("../../data/Human_DLPFC_Visium_processedData_sce_scran.Rdata")
sce

gene_names <- rowData(sce)$gene_name
names(gene_names) <- rowData(sce)$gene_id

# checks
head(gene_names)
length(gene_names)
length(unique(gene_names))

length(unique(unlist(spatialDE_sig_all)))
length(unique(unlist(spatialDE_sig_sub)))

length(unique(unlist(spatialDE_zero_all)))
length(unique(unlist(spatialDE_zero_sub)))

stopifnot(all(unique(unlist(spatialDE_sig_all)) %in% unique(names(gene_names))))
table(unique(unlist(spatialDE_sig_all)) %in% unique(names(gene_names)))
stopifnot(all(unique(unlist(spatialDE_sig_sub)) %in% unique(names(gene_names))))
table(unique(unlist(spatialDE_sig_sub)) %in% unique(names(gene_names)))

stopifnot(all(unique(unlist(spatialDE_zero_all)) %in% unique(names(gene_names))))
table(unique(unlist(spatialDE_zero_all)) %in% unique(names(gene_names)))
stopifnot(all(unique(unlist(spatialDE_zero_sub)) %in% unique(names(gene_names))))
table(unique(unlist(spatialDE_zero_sub)) %in% unique(names(gene_names)))

# convert gene symbols to gene names
spatialDE_sig_all <- lapply(spatialDE_sig_all, function(d) {
    unname(gene_names[d])
})
spatialDE_sig_sub <- lapply(spatialDE_sig_sub, function(d) {
    unname(gene_names[d])
})

spatialDE_zero_all <- lapply(spatialDE_zero_all, function(d) {
    unname(gene_names[d])
})
spatialDE_zero_sub <- lapply(spatialDE_zero_sub, function(d) {
    unname(gene_names[d])
})
```


```{r}
# pooled gene lists
spatialDE_sig_all_pool <- sort(unique(unlist(spatialDE_sig_all)))
spatialDE_sig_sub_pool <- sort(unique(unlist(spatialDE_sig_sub)))

spatialDE_zero_all_pool <- sort(unique(unlist(spatialDE_zero_all)))
spatialDE_zero_sub_pool <- sort(unique(unlist(spatialDE_zero_sub)))

length(spatialDE_sig_all_pool)
length(spatialDE_sig_sub_pool)

length(spatialDE_zero_all_pool)
length(spatialDE_zero_sub_pool)
```


## Plots

Plot the following:

- sample vs. sample: all spots, significant genes (q-values < 0.05)
- sample vs. sample: all spots, genes with q-values = 0
- sample vs. sample: subsampled spots, significant genes (q-values < 0.05)
- sample vs. sample: subsampled spots, genes with q-values = 0
- each sample (all spots vs. subsampling): significant genes (q-values < 0.05)
- each sample (all spots vs. subsampling): genes with q-values = 0
- pooled lists: significant genes (q-values < 0.05)
- pooled lists: genes with q-values = 0

```{r, fig.width = 12, fig.height = 10}
# UpSetR plots

# -----------------
# sample vs. sample
# -----------------

# all spots, significant genes (q-values < 0.05)
#pdf("../plots/intersections/upset_SpatialDE_all_sig.pdf", width = 12, height = 10)
upset(
    fromList(spatialDE_sig_all), 
    nsets = length(spatialDE_sig_all), 
    nintersects = NA, 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(sample_names), 
    keep.order = TRUE, 
    mainbar.y.label = "SpatialDE intersections: all spots, significant genes (q-values < 0.05)"
)
#dev.off()

# all spots, genes with q-values = 0
#pdf("../plots/intersections/upset_SpatialDE_all_zero.pdf", width = 12, height = 10)
upset(
    fromList(spatialDE_zero_all), 
    nsets = length(spatialDE_zero_all), 
    nintersects = NA, 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(sample_names), 
    keep.order = TRUE, 
    mainbar.y.label = "SpatialDE intersections: all spots, genes with q-values = 0"
)
#dev.off()

# subsampled spots, significant genes (q-values < 0.05)
#pdf("../plots/intersections/upset_SpatialDE_sub_sig.pdf", width = 12, height = 10)
upset(
    fromList(spatialDE_sig_sub), 
    nsets = length(spatialDE_sig_sub), 
    nintersects = NA, 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(sample_names), 
    keep.order = TRUE, 
    mainbar.y.label = "SpatialDE intersections: subsampled spots, significant genes (q-values < 0.05)"
)
#dev.off()

# subsampled spots, genes with q-values = 0
#pdf("../plots/intersections/upset_SpatialDE_sub_zero.pdf", width = 12, height = 10)
upset(
    fromList(spatialDE_zero_sub), 
    nsets = length(spatialDE_zero_sub), 
    nintersects = NA, 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(sample_names), 
    keep.order = TRUE, 
    mainbar.y.label = "SpatialDE intersections: subsampled spots, genes with q-values = 0"
)
#dev.off()
```


```{r, fig.width = 4, fig.height = 4}
# UpSetR plots

# ---------------------------------------
# each sample (all spots vs. subsampling)
# ---------------------------------------

# significant genes (q-values < 0.05)
for (i in 1:length(sample_names)) {
    fn <- paste0("../plots/intersections/upset_SpatialDE_subsampling_comparison_", gsub("sample_", "", sample_names[i]), "_sig.pdf")
    #pdf(fn, width = 4, height = 4)
    print(upset(
        fromList(list(all_spots = spatialDE_sig_all[[i]], subsampled = spatialDE_sig_sub[[i]])), 
        order.by = "degree", 
        decreasing = TRUE, 
        sets = rev(c("all_spots", "subsampled")), 
        keep.order = TRUE, 
        mainbar.y.label = sample_names[i]
    ))
    #dev.off()
}

# genes with q-values = 0
for (i in 1:length(sample_names)) {
    fn <- paste0("../plots/intersections/upset_SpatialDE_subsampling_comparison_", gsub("sample_", "", sample_names[i]), "_zero.pdf")
    #pdf(fn, width = 4, height = 4)
    print(upset(
        fromList(list(all_spots = spatialDE_zero_all[[i]], subsampled = spatialDE_zero_sub[[i]])), 
        order.by = "degree", 
        decreasing = TRUE, 
        sets = rev(c("all_spots", "subsampled")), 
        keep.order = TRUE, 
        mainbar.y.label = sample_names[i]
    ))
    #dev.off()
}
```


```{r, fig.width = 5, fig.height = 6}
# UpSetR plots

# ------------
# pooled lists
# ------------

# pooled samples, significant genes (q-values < 0.05)
#pdf("../plots/intersections/upset_SpatialDE_pooled_sig.pdf", width = 5, height = 6)
upset(
    fromList(list(all_spots = spatialDE_sig_all_pool, subsampled = spatialDE_sig_sub_pool)), 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(c("all_spots", "subsampled")), 
    keep.order = TRUE, 
    mainbar.y.label = "SpatialDE intersections: pooled samples, significant genes (q-values < 0.05)"
)
#dev.off()

# pooled samples, genes with q-values = 0
#pdf("../plots/intersections/upset_SpatialDE_pooled_zero.pdf", width = 5, height = 6)
upset(
    fromList(list(all_spots = spatialDE_zero_all_pool, subsampled = spatialDE_zero_sub_pool)), 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(c("all_spots", "subsampled")), 
    keep.order = TRUE, 
    mainbar.y.label = "SpatialDE intersections: pooled samples, genes with q-values = 0"
)
#dev.off()
```



