---
title: "SpatialDE subsampling comparison"
author: "Lukas Weber"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
    html_document:
        toc: true
        toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# SpatialDE subsampling comparison

Comparison of gene lists generated from Stephanie Hicks's SpatialDE runs with and without subsampling on the number of spots. (The version with subsampling used subsampling to 1500 spots per sample for faster runtime.)


## Load data

Load two sets of SpatialDE results for each sample:

- all spots (no subsampling)
- subsampling to 1500 spots per sample


```{r, message = FALSE}
library(SingleCellExperiment)
library(readr)
library(dplyr)
library(UpSetR)
```


```{r, message = FALSE, warning = FALSE}
# load spreadsheets containing SpatialDE results
spatialDE_files_all <- list.files(
    "../../data/SpatialDE/spatialDE_all", pattern = "\\.csv$", full.names = TRUE
)
spatialDE_files_sub <- list.files(
    "../../data/SpatialDE/spatialDE_subsampled", pattern = "\\.csv$", full.names = TRUE
)

spatialDE_results_all <- lapply(spatialDE_files_all, read_csv)
spatialDE_results_sub <- lapply(spatialDE_files_sub, read_csv)

sample_names_all <- gsub("_spatialDE.*$", "", gsub("^.*/", "", spatialDE_files_all))
sample_names_sub <- gsub("_spatialDE.*$", "", gsub("^.*/", "", spatialDE_files_sub))

stopifnot(all(sample_names_all == sample_names_sub))

sample_names <- sample_names_all

names(spatialDE_results_all) <- sample_names
names(spatialDE_results_sub) <- sample_names
```


## Comparison of SpatialDE gene lists

Compare the following:

- total number of genes per sample
- number of significant genes per sample (q-values < 0.05)

```{r}
# check numbers of genes

# total number of genes per sample (note numbers differ slightly)
sapply(spatialDE_results_all, nrow)
sapply(spatialDE_results_sub, nrow)

# number of significant genes per sample (q-values < 0.05)
sapply(spatialDE_results_all, function(d) table(d$qval < 0.05))
sapply(spatialDE_results_sub, function(d) table(d$qval < 0.05))
```


```{r}
# extract gene lists

# significant genes per sample (q-values < 0.05)
spatialDE_sig_all <- lapply(spatialDE_results_all, function(d) {
    filter(d, qval < 0.05)$g
})
spatialDE_sig_sub <- lapply(spatialDE_results_sub, function(d) {
    filter(d, qval < 0.05)$g
})
```


```{r}
# convert gene symbols to gene names

# load scran output file from Leo
load("../../data/Human_DLPFC_Visium_processedData_sce_scran.Rdata")
sce

gene_names <- rowData(sce)$gene_name
names(gene_names) <- rowData(sce)$gene_id

# checks
head(gene_names)
length(gene_names)
length(unique(gene_names))

length(unique(unlist(spatialDE_sig_all)))
length(unique(unlist(spatialDE_sig_sub)))

stopifnot(all(unique(unlist(spatialDE_sig_all)) %in% unique(names(gene_names))))
table(unique(unlist(spatialDE_sig_all)) %in% unique(names(gene_names)))
stopifnot(all(unique(unlist(spatialDE_sig_sub)) %in% unique(names(gene_names))))
table(unique(unlist(spatialDE_sig_sub)) %in% unique(names(gene_names)))

# convert gene symbols to gene names
spatialDE_sig_all <- lapply(spatialDE_sig_all, function(d) {
    data.frame(gene_symbol = d, gene_name = gene_names[d])
})
spatialDE_sig_sub <- lapply(spatialDE_sig_sub, function(d) {
    data.frame(gene_symbol = d, gene_name = gene_names[d])
})
```


```{r}
# save spreadsheets
for (i in seq_along(sample_names)) {
    write_csv(
        as.data.frame(spatialDE_sig_all[[i]]), 
        paste0("../outputs/genes/SpatialDE_genes_sig_all_", gsub("sample_", "", sample_names[i]), ".csv")
    )
    write_csv(
        as.data.frame(spatialDE_sig_sub[[i]]), 
        paste0("../outputs/genes/SpatialDE_genes_sig_sub_", gsub("sample_", "", sample_names[i]), ".csv")
    )
}
```


```{r}
# pooled gene lists
spatialDE_sig_all_pool <- unique(do.call("rbind", spatialDE_sig_all))
spatialDE_sig_sub_pool <- unique(do.call("rbind", spatialDE_sig_sub))

nrow(spatialDE_sig_all_pool)
nrow(spatialDE_sig_sub_pool)

# save spreadsheets
write_csv(
    as.data.frame(spatialDE_sig_all_pool), 
    "../outputs/genes/SpatialDE_genes_sig_all_pooled.csv"
)
write_csv(
    as.data.frame(spatialDE_sig_sub_pool), 
    "../outputs/genes/SpatialDE_genes_sig_sub_pooled.csv"
)
```


## Plots

Plot the following:

- sample vs. sample: all spots, significant genes (q-values < 0.05)
- sample vs. sample: subsampled spots, significant genes (q-values < 0.05)
- each sample (all spots vs. subsampling): significant genes (q-values < 0.05)
- pooled lists: significant genes (q-values < 0.05)

```{r, fig.width = 12, fig.height = 10}
# UpSetR plots

# -----------------
# sample vs. sample
# -----------------

# all spots, significant genes (q-values < 0.05)
#pdf("../plots/SpatialDE_subsampling_comparison/upset_SpatialDE_all_sig.pdf", width = 12, height = 10)
upset(
    fromList(lapply(spatialDE_sig_all, function(d) as.character(d[, "gene_symbol"]))), 
    nsets = length(spatialDE_sig_all), 
    nintersects = NA, 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(sample_names), 
    keep.order = TRUE, 
    mainbar.y.label = "SpatialDE intersections: all spots, significant genes (q-values < 0.05)"
)
#dev.off()

# subsampled spots, significant genes (q-values < 0.05)
#pdf("../plots/SpatialDE_subsampling_comparison/upset_SpatialDE_sub_sig.pdf", width = 12, height = 10)
upset(
    fromList(lapply(spatialDE_sig_sub, function(d) as.character(d[, "gene_symbol"]))), 
    nsets = length(spatialDE_sig_sub), 
    nintersects = NA, 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(sample_names), 
    keep.order = TRUE, 
    mainbar.y.label = "SpatialDE intersections: subsampled spots, significant genes (q-values < 0.05)"
)
#dev.off()
```


```{r, fig.width = 4, fig.height = 4}
# UpSetR plots

# ---------------------------------------
# each sample (all spots vs. subsampling)
# ---------------------------------------

# significant genes (q-values < 0.05)
for (i in 1:length(sample_names)) {
    fn <- paste0("../plots/SpatialDE_subsampling_comparison/upset_SpatialDE_subsampling_comparison_", 
                 gsub("sample_", "", sample_names[i]), "_sig.pdf")
    #pdf(fn, width = 4, height = 4)
    print(upset(
        fromList(list(all_spots = as.character(spatialDE_sig_all[[i]][, "gene_symbol"]), 
                      subsampled = as.character(spatialDE_sig_sub[[i]][, "gene_symbol"]))), 
        order.by = "degree", 
        decreasing = TRUE, 
        sets = rev(c("all_spots", "subsampled")), 
        keep.order = TRUE, 
        mainbar.y.label = sample_names[i]
    ))
    #dev.off()
}
```


```{r, fig.width = 5, fig.height = 6}
# UpSetR plots

# ------------
# pooled lists
# ------------

# pooled samples, significant genes (q-values < 0.05)
#pdf("../plots/SpatialDE_subsampling_comparison/upset_SpatialDE_pooled_sig.pdf", width = 5, height = 6)
upset(
    fromList(list(all_spots = as.character(spatialDE_sig_all_pool[, "gene_symbol"]), 
                  subsampled = as.character(spatialDE_sig_sub_pool[, "gene_symbol"]))), 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(c("all_spots", "subsampled")), 
    keep.order = TRUE, 
    mainbar.y.label = "SpatialDE intersections: pooled samples, significant genes (q-values < 0.05)"
)
#dev.off()
```



