---
title: "SpatialDE genes analysis"
author: "Lukas Weber"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
    html_document:
        toc: true
        toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# SpatialDE genes analysis

Analysis of SpatialDE gene lists and overlap with other gene lists. Note the SpatialDE gene lists (before filtering) are saved as .csv spreadsheet outputs from the script "SpatialDE_subsampling_comparison.Rmd".

```{r, message = FALSE}
library(SingleCellExperiment)
library(readr)
library(readxl)
library(dplyr)
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(UpSetR)
library(patchwork)
```


## Load gene lists: SpatialDE

```{r}
# load spreadsheets containing SpatialDE gene lists
genes_SpatialDE_files <- list.files(
    "../outputs/SpatialDE_subsampling_comparison", 
    pattern = "^SpatialDE_genes_sig_all_[0-9]+\\.csv$", full.names = TRUE
)
genes_SpatialDE_files_pooled <- list.files(
    "../outputs/SpatialDE_subsampling_comparison", 
    pattern = "^SpatialDE_genes_sig_all_pooled\\.csv$", full.names = TRUE
)

genes_SpatialDE_files
genes_SpatialDE_files_pooled

genes_SpatialDE <- lapply(genes_SpatialDE_files, read_csv)

genes_SpatialDE_pooled <- read_csv(genes_SpatialDE_files_pooled)

sample_names <- paste0("sample_", gsub("\\.csv$", "", gsub("^.*_", "", genes_SpatialDE_files)))
sample_names

names(genes_SpatialDE) <- sample_names

# number of genes
sapply(genes_SpatialDE, nrow)
nrow(genes_SpatialDE_pooled)
```


## Load gene lists: top HVGs

```{r}
# load scran output file from Leo
load("../../data/Human_DLPFC_Visium_processedData_sce_scran.Rdata")
sce

# top HVGs
head(top.hvgs)
length(top.hvgs)

genes_HVGs <- rowData(sce)[top.hvgs, c("gene_id", "gene_name")]

# number of genes
head(genes_HVGs)
dim(genes_HVGs)
```


## Load gene lists: pseudobulk layers (from Leo's analyses)

```{r}
# load spreadsheet of significant genes for pseudobulk layers (from Leo's analyses)
file_genes_pseudobulk <- "Layer_Guesses/sig_genes.csv"
sig_genes <- read_csv(file_genes_pseudobulk)

sig_genes

genes_pseudobulk <- sig_genes[, c("ensembl", "gene")]
colnames(genes_pseudobulk) <- c("gene_id", "gene_name")

dim(genes_pseudobulk)

# remove duplicates (i.e. genes identified as significant for multiple layers)
genes_pseudobulk <- distinct(genes_pseudobulk)

dim(genes_pseudobulk)
```


## Filtering: SpatialDE genes

SpatialDE detects a large number of significant (q-values < 0.05) genes per sample, some of which may not be particularly meaningful for defining cell types associated with cortex layers (e.g. due to very low expression, or mitochondrial genes). For our subsequent analyses, we filter the SpatialDE gene lists as follows:

- remove genes with very low UMI counts (e.g. total of 1000 counts summed across spots per sample)
- remove mitochondrial genes (gene names starting with "MT-")

```{r}
# parameters
n_UMI_filt <- 1000

sce_nofilt_list <- list()
sce_filt_list <- list()
genes_SpatialDE_filt <- list()

# for each sample
for (i in seq_along(sample_names)) {
    
    # select spots from this sample
    sce_sub <- sce[, colData(sce)$sample_name == gsub("^sample_", "", sample_names[i])]
    dim(sce_sub)
    
    # subset genes from SpatialDE
    sce_sub_SpatialDE <- sce_sub[genes_SpatialDE[[i]]$gene_id, ]
    dim(sce_sub_SpatialDE)
    nrow(sce_sub_SpatialDE) == nrow(genes_SpatialDE[[i]])
    
    sce_nofilt_list[[i]] <- sce_sub_SpatialDE
    
    # filtering: remove genes with very low total UMI counts (summed across spots)
    ix_low_UMI <- rowSums(counts(sce_sub_SpatialDE)) < n_UMI_filt
    # note: this filters out a large proportion of genes
    table(ix_low_UMI)
    
    # filtering: remove mitochondrial genes
    ix_mito <- grepl("^MT-", toupper(rowData(sce_sub_SpatialDE)$gene_name))
    table(ix_mito)
    rowData(sce)$gene_name[ix_mito]
    
    stopifnot(length(ix_low_UMI) == nrow(sce_sub_SpatialDE))
    stopifnot(length(ix_mito) == nrow(sce_sub_SpatialDE))
    
    # apply filtering
    ix_keep <- (!ix_low_UMI) & (!ix_mito)
    
    print(sample_names[i])
    print(table(ix_keep))
    
    sce_filt <- sce_sub_SpatialDE[ix_keep, ]
    dim(sce_filt)
    
    sce_filt_list[[i]] <- sce_filt
    
    genes_SpatialDE_filt_sub <- as.data.frame(rowData(sce_filt)[, c("gene_id", "gene_name")])
    rownames(genes_SpatialDE_filt_sub) <- NULL
    
    genes_SpatialDE_filt[[i]] <- genes_SpatialDE_filt_sub
    
    # save spreadsheets
    file_out <- paste0("../outputs/SpatialDE_genes_analysis/SpatialDE_filtered_", sample_names[i], ".csv")
    write_csv(genes_SpatialDE_filt_sub, file_out)
}


names(sce_filt_list) <- sample_names
names(genes_SpatialDE_filt) <- sample_names


# pooled gene list (no filtering)
genes_SpatialDE_pooled <- unique(do.call("rbind", genes_SpatialDE))
rownames(genes_SpatialDE_pooled) <- NULL
dim(genes_SpatialDE_pooled)

# pooled gene list (with additional filtering)
genes_SpatialDE_filt_pooled <- unique(do.call("rbind", genes_SpatialDE_filt))
rownames(genes_SpatialDE_filt_pooled) <- NULL
dim(genes_SpatialDE_filt_pooled)


# save spreadsheet
file_out <- "../outputs/SpatialDE_genes_analysis/SpatialDE_filtered_pooled.csv"
write_csv(genes_SpatialDE_filt_pooled, file_out)
```


## Export supplementary table

Export supplementary table showing number of SpatialDE genes per sample (before and after filtering).

```{r}
# supplementary table showing number of SpatialDE genes per sample

supp_table <- data.frame(
    sample = gsub("^sample_", "", sample_names), 
    n_SpatialDE_genes_before_filt = sapply(genes_SpatialDE, nrow), 
    n_SpatialDE_genes_after_filt = sapply(genes_SpatialDE_filt, nrow), 
    stringsAsFactors = FALSE
)

colnames(supp_table) <- 
    c("Sample", 
      "Number of significant SpatialDE genes (before filtering)", 
      "Number of significant SpatialDE genes (after filtering)")

supp_table <- rbind(
    supp_table, 
    c("Total number of unique genes", 
      nrow(genes_SpatialDE_pooled), 
      nrow(genes_SpatialDE_filt_pooled)))

rownames(supp_table) <- NULL

write.csv(supp_table, "../outputs/supp_tables/SuppTable_SpatialDE_genes.csv", row.names = FALSE)
```


## Plots: gene sets

Plots showing the following:

- overlap between gene sets (using pooled data from all samples combined): SpatialDE (no filtering), SpatialDE (with additional filtering), HVGs, pseudobulk layers
- number of SpatialDE genes per sample

```{r, fig.width = 6.5, fig.height = 6}
# plot showing overlap between gene sets (using pooled lists)
list_overlap <- list(
    SpatialDE_pooled = genes_SpatialDE_pooled$gene_id, 
    SpatialDE_pooled_filtered = genes_SpatialDE_filt_pooled$gene_id, 
    HVGs = genes_HVGs$gene_id, 
    pseudobulk_layers = genes_pseudobulk$gene_id
)

#pdf("../plots/SpatialDE_genes_analysis/upset_genes_SpatialDE_SpatialDEfilt_HVGs_pseudobulk.pdf", width = 6.5, height = 6)
upset(
    fromList(list_overlap), 
    nsets = length(list_overlap), 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(names(list_overlap)), 
    set_size.show = TRUE, 
    set_size.scale_max = 8000, 
    keep.order = TRUE
)
#dev.off()


# plot showing overlap between gene sets (using pooled lists)
list_overlap <- list(
    SpatialDE_pooled_filtered = genes_SpatialDE_filt_pooled$gene_id, 
    HVGs = genes_HVGs$gene_id, 
    pseudobulk_layers = genes_pseudobulk$gene_id
)

#pdf("../plots/SpatialDE_genes_analysis/upset_genes_SpatialDEfilt_HVGs_pseudobulk.pdf", width = 6.5, height = 6)
upset(
    fromList(list_overlap), 
    nsets = length(list_overlap), 
    order.by = "degree", 
    decreasing = TRUE, 
    sets = rev(names(list_overlap)), 
    set_size.show = TRUE, 
    set_size.scale_max = 4000, 
    keep.order = TRUE
)
#dev.off()
```


```{r, fig.width = 6.5, fig.height = 5}
# plot showing number of significant SpatialDE genes per sample (no filtering)
n_genes_SpatialDE <- sapply(genes_SpatialDE, nrow)
n_genes_SpatialDE

names(n_genes_SpatialDE) <- gsub("^sample_", "", names(n_genes_SpatialDE))

d_plot <- data.frame(
    sample = names(n_genes_SpatialDE), 
    n_genes = unname(n_genes_SpatialDE)
)

ggplot(d_plot, aes(x = sample, y = n_genes)) + 
    geom_bar(stat = "identity") + 
    geom_text(aes(label = n_genes), vjust = -0.5, size = 3) + 
    ggtitle("Number of significant SpatialDE genes per sample (no filtering)") + 
    ylab("Number of significant genes (q-values < 0.05)") + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5), 
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank())

filename <- "../plots/SpatialDE_genes_analysis/n_genes_SpatialDE.pdf"
ggsave(filename, width = 6.5, height = 5)


# plot showing number of significant SpatialDE genes per sample (with additional filtering)
n_genes_SpatialDE_filt <- sapply(genes_SpatialDE_filt, nrow)
n_genes_SpatialDE_filt

names(n_genes_SpatialDE_filt) <- gsub("^sample_", "", names(n_genes_SpatialDE_filt))

d_plot <- data.frame(
    sample = names(n_genes_SpatialDE_filt), 
    n_genes = unname(n_genes_SpatialDE_filt)
)

ggplot(d_plot, aes(x = sample, y = n_genes)) + 
    geom_bar(stat = "identity") + 
    geom_text(aes(label = n_genes), vjust = -0.5, size = 3) + 
    ggtitle("Number of significant SpatialDE genes per sample (filtered)") + 
    ylab("Number of significant genes (q-values < 0.05)") + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5), 
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank())

filename <- "../plots/SpatialDE_genes_analysis/n_genes_SpatialDE_filtered.pdf"
ggsave(filename, width = 6.5, height = 5)
```


## Plots: individual genes

Plots of spatial expression patterns for individual genes from SpatialDE.

These plots can be used to identify marker genes for layers, or marker genes for unexpected spatial structures (e.g. blood vessels).

Note: we only show plots for sample 151673, which had the cleanest layer structure. After filtering, there are 1950 significant SpatialDE genes (q-values < 0.05) for sample 151673.

```{r, fig.width = 12, fig.height = 20}
# select sample
i <- 9
sce_filt <- sce_filt_list[[i]]
dim(sce_filt)

# extract x-y coordinates of spots (note: y coordinate is reversed)
xy_coords <- data.frame(
    x_coord = colData(sce_filt)[, c("imagecol")], 
    y_coord = -colData(sce_filt)[, c("imagerow")]
)

# extract expression levels (UMI counts)
exprs_filt <- counts(sce_filt)
dim(exprs_filt)
# replace gene IDs with gene names for plots
stopifnot(all(rownames(exprs_filt) == rowData(sce_filt)$gene_id))
rownames(exprs_filt) <- rowData(sce_filt)$gene_name
# check column ordering matches spots
stopifnot(all(colData(sce_filt)$barcode == colnames(exprs_filt)))
stopifnot(length(colData(sce_filt)$barcode) == length(colnames(exprs_filt)))
# order genes alphabetically
exprs_filt <- exprs_filt[order(rownames(exprs_filt)), ]

d_plot <- cbind(
    barcode = colData(sce_filt)$barcode, 
    xy_coords, 
    as.data.frame(as.matrix(t(exprs_filt)))
)

d_plot <- melt(
    d_plot, 
    id.vars = c("barcode", "x_coord", "y_coord"), 
    variable.name = "gene_name", 
    value.name = "UMIs"
)

max_UMI <- max(d_plot$UMIs)

# split into subplots since otherwise too many panels and code too slow
# up to n_subplots depending on number of panels
n_subplots <- 10
n_col <- 14
n_row <- 18
d_plot$subplot <- rep(1:n_subplots, each = n_col * n_row * ncol(sce_filt))[1:(nrow(sce_filt) * ncol(sce_filt))]

for (z in unique(d_plot$subplot)) {
    d_plot_sub <- d_plot[d_plot$subplot == z, ]
    print(
        ggplot(d_plot_sub, aes(x = x_coord, y = y_coord, color = UMIs)) + 
            facet_wrap(~ gene_name, ncol = n_col) + 
            geom_point(size = 0.1, alpha = 0.25) + 
            scale_color_gradientn(colors = c("gray95", "deepskyblue", "blue", "black"), 
                                  values = rescale(c(0, 3, 15, max_UMI)), 
                                  limits = c(0, max_UMI)) + 
            coord_fixed() + 
            ggtitle(paste0("UMIs of significant SpatialDE genes (filtered): sample ", 
                           gsub("^sample_", "", sample_names[i]))) + 
            theme_bw() + 
            theme(strip.text.x = element_text(size = 5), 
                  axis.text.x = element_text(size = 5), 
                  axis.text.y = element_text(size = 5), 
                  panel.grid.major = element_blank(), 
                  panel.grid.minor = element_blank())
    )
    
    filename <- paste0("../plots/SpatialDE_genes_analysis/SpatialDE_genes_expr_sample_", 
                       gsub("^sample_", "", sample_names[i]), "_part_", z, ".png")
    ggsave(filename, width = 12, height = 20)
}
```


## Comparison with known marker genes

Compare gene lists with known marker genes from Kristen Maynard (saved in spreadsheet `KRM_Layer_Markers.xlsx`).

```{r}
# -------------------------------
# comparison of marker gene lists
# -------------------------------

# load names of marker genes
KRM_Layer_Markers <- read_xlsx("KRM_Layer_Markers.xlsx")
KRM_Layer_Markers
dim(KRM_Layer_Markers)

marker_genes_KRM <- KRM_Layer_Markers$Gene


# for each sample
for (i in seq_along(sample_names)) {
    
    # overlap with SpatialDE significant genes (no filtering)
    sum(toupper(marker_genes_KRM) %in% toupper(rowData(sce_nofilt_list[[i]])$gene_name))
    sum(toupper(rowData(sce_nofilt_list[[i]])$gene_name) %in% toupper(marker_genes_KRM))
    in_SpatialDE_nofilt <- toupper(marker_genes_KRM) %in% toupper(rowData(sce_nofilt_list[[i]])$gene_name)
    
    # overlap with SpatialDE significant genes (with additional filtering)
    sum(toupper(marker_genes_KRM) %in% toupper(rowData(sce_filt_list[[i]])$gene_name))
    sum(toupper(rowData(sce_filt_list[[i]])$gene_name) %in% toupper(marker_genes_KRM))
    in_SpatialDE_filt <- toupper(marker_genes_KRM) %in% toupper(rowData(sce_filt_list[[i]])$gene_name)
    
    # overlap with SpatialDE significant genes: pooled samples (no filtering)
    sum(toupper(marker_genes_KRM) %in% toupper(genes_SpatialDE_pooled$gene_name))
    sum(toupper(genes_SpatialDE_pooled$gene_name) %in% toupper(marker_genes_KRM))
    in_SpatialDE_nofilt_pooled <- toupper(marker_genes_KRM) %in% toupper(genes_SpatialDE_pooled$gene_name)
    
    # overlap with SpatialDE significant genes: pooled samples (no filtering)
    sum(toupper(marker_genes_KRM) %in% toupper(genes_SpatialDE_filt_pooled$gene_name))
    sum(toupper(genes_SpatialDE_filt_pooled$gene_name) %in% toupper(marker_genes_KRM))
    in_SpatialDE_filt_pooled <- toupper(marker_genes_KRM) %in% toupper(genes_SpatialDE_filt_pooled$gene_name)
    
    # overlap with full list of HVGs
    sum(toupper(marker_genes_KRM) %in% toupper(genes_HVGs$gene_name))
    sum(toupper(genes_HVGs$gene_name) %in% toupper(marker_genes_KRM))
    in_HVGs <- toupper(marker_genes_KRM) %in% toupper(genes_HVGs$gene_name)
    
    stopifnot(length(marker_genes_KRM) == length(in_SpatialDE_nofilt))
    stopifnot(length(marker_genes_KRM) == length(in_SpatialDE_filt))
    stopifnot(length(marker_genes_KRM) == length(in_SpatialDE_nofilt_pooled))
    stopifnot(length(marker_genes_KRM) == length(in_SpatialDE_filt_pooled))
    stopifnot(length(marker_genes_KRM) == length(in_HVGs))
    
    # save spreadsheet
    KRM_Layer_Markers_overlap <- cbind(
        KRM_Layer_Markers, 
        in_SpatialDE_nofilt, 
        in_SpatialDE_filt, 
        in_SpatialDE_nofilt_pooled, 
        in_SpatialDE_filt_pooled, 
        in_HVGs
    )
    
    filename <- paste0("../outputs/SpatialDE_genes_analysis/KRM_Layer_Markers_overlap_", sample_names[i], ".csv")
    write_csv(KRM_Layer_Markers_overlap, filename)
}
```


```{r, fig.width = 9, fig.height = 4}
# ----------------------------------------------------
# plots of 3 known marker genes (from Kristen Maynard)
# ----------------------------------------------------

# for each sample
for (i in seq_along(sample_names)) {
    
    # plots of 3 known marker genes (MOBP, PCP4, SNAP2)
    marker_genes_3 <- c("MOBP", "PCP4", "SNAP25")
    
    # subset original SCE object
    sce_3 <- sce[toupper(rowData(sce)$gene_name) %in% marker_genes_3, 
                 colData(sce)$sample_name == gsub("^sample_", "", sample_names[i])]
    dim(sce_3)
    
    # extract x-y coordinates of spots (note: y coordinate is reversed)
    xy_coords <- data.frame(
        x_coord = colData(sce_3)[, c("imagecol")], 
        y_coord = -colData(sce_3)[, c("imagerow")]
    )
    
    # extract expression levels (UMI counts)
    exprs_3 <- counts(sce_3)
    dim(exprs_3)
    # replace gene IDs with gene names for plots
    stopifnot(all(rownames(exprs_3) == rowData(sce_3)$gene_id))
    rownames(exprs_3) <- rowData(sce_3)$gene_name
    # check column ordering matches spots
    stopifnot(all(colData(sce_3)$barcode == colnames(exprs_3)))
    stopifnot(length(colData(sce_3)$barcode) == length(colnames(exprs_3)))
    # order genes alphabetically
    exprs_3 <- exprs_3[order(rownames(exprs_3)), ]
    
    d_plot <- cbind(
        barcode = colData(sce_3)$barcode, 
        xy_coords, 
        as.data.frame(as.matrix(t(exprs_3)))
    )
    
    d_plot <- melt(
        d_plot, 
        id.vars = c("barcode", "x_coord", "y_coord"), 
        variable.name = "gene_name", 
        value.name = "UMIs"
    )
    
    max_UMI <- max(d_plot$UMIs)
    
    print(
        ggplot(d_plot, aes(x = x_coord, y = y_coord, color = UMIs)) + 
            facet_wrap(~ gene_name) + 
            geom_point(size = 0.6) + 
            scale_color_gradientn(colors = c("gray95", "deepskyblue", "blue", "black"), 
                                  values = rescale(c(0, 3, 15, max_UMI)), 
                                  limits = c(0, max_UMI)) + 
            coord_fixed() + 
            ggtitle(paste0("UMIs of known marker genes: sample ", gsub("^sample_", "", sample_names[i]))) + 
            theme_bw() + 
            theme(panel.grid.major = element_blank(), 
                  panel.grid.minor = element_blank())
    )
    
    filename <- paste0("../plots/SpatialDE_genes_analysis/marker_genes_3_expr_sample_", 
                       gsub("^sample_", "", sample_names[i]), ".png")
    ggsave(filename, width = 9, height = 4)
}
```


```{r, fig.width = 8, fig.height = 10}
# ------------------------------------------------------
# plots of all known marker genes (from Kristen Maynard)
# ------------------------------------------------------

# for each sample
for (i in seq_along(sample_names)) {
    
    length(marker_genes_KRM)
    
    # subset original SCE object
    sce_KRM <- sce[toupper(rowData(sce)$gene_name) %in% toupper(marker_genes_KRM), 
                   colData(sce)$sample_name == gsub("^sample_", "", sample_names[i])]
    # note: not all marker genes are in SCE object
    dim(sce_KRM)
    
    # extract x-y coordinates of spots (note: y coordinate is reversed)
    xy_coords <- data.frame(
        x_coord = colData(sce_KRM)[, c("imagecol")], 
        y_coord = -colData(sce_KRM)[, c("imagerow")]
    )
    
    # extract expression levels (UMI counts)
    exprs_KRM <- counts(sce_KRM)
    dim(exprs_KRM)
    # replace gene IDs with gene names for plots
    stopifnot(all(rownames(exprs_KRM) == rowData(sce_KRM)$gene_id))
    rownames(exprs_KRM) <- rowData(sce_KRM)$gene_name
    # check column ordering matches spots
    stopifnot(all(colData(sce_KRM)$barcode == colnames(exprs_KRM)))
    stopifnot(length(colData(sce_KRM)$barcode) == length(colnames(exprs_KRM)))
    # order genes alphabetically
    exprs_KRM <- exprs_KRM[order(rownames(exprs_KRM)), ]
    
    d_plot <- cbind(
        barcode = colData(sce_KRM)$barcode, 
        xy_coords, 
        as.data.frame(as.matrix(t(exprs_KRM)))
    )
    
    d_plot <- melt(
        d_plot, 
        id.vars = c("barcode", "x_coord", "y_coord"), 
        variable.name = "gene_name", 
        value.name = "UMIs"
    )
    
    max_UMI <- max(d_plot$UMIs)
    
    print(
        ggplot(d_plot, aes(x = x_coord, y = y_coord, color = UMIs)) + 
            facet_wrap(~ gene_name) + 
            geom_point(size = 0.1, alpha = 0.25) + 
            scale_color_gradientn(colors = c("gray95", "deepskyblue", "blue", "black"), 
                                  values = rescale(c(0, 3, 15, max_UMI)), 
                                  limits = c(0, max_UMI)) + 
            coord_fixed() + 
            ggtitle(paste0("UMIs of known marker genes: sample ", gsub("^sample_", "", sample_names[i]))) + 
            theme_bw() + 
            theme(strip.text.x = element_text(size = 7), 
                  axis.text.x = element_text(size = 7), 
                  axis.text.y = element_text(size = 7), 
                  panel.grid.major = element_blank(), 
                  panel.grid.minor = element_blank())
    )
    
    filename <- paste0("../plots/SpatialDE_genes_analysis/marker_genes_KRM_expr_sample_", 
                       gsub("^sample_", "", sample_names[i]), ".png")
    ggsave(filename, width = 8, height = 10)
}
```


## Plots: figure

Plots for final figure:

- intersections of gene lists
- expression plot for gene HBB (sample 151673)

```{r, fig.width = 6, fig.height = 5}
# get gene IDs for KRM markers
marker_genes_KRM_IDs <- rowData(sce)$gene_id[match(toupper(marker_genes_KRM), rowData(sce)$gene_name)]
marker_genes_KRM_IDs[is.na(marker_genes_KRM_IDs)] <- toupper(marker_genes_KRM)[is.na(marker_genes_KRM_IDs)]

# upset plot for final figure
# note: using filtered SpatialDE lists
list_overlap <- list(
    SpatialDE_pooled = genes_SpatialDE_filt_pooled$gene_id, 
    SpatialDE_sample_151673 = genes_SpatialDE_filt[[9]]$gene_id, 
    HVGs = genes_HVGs$gene_id, 
    pseudobulk_layers = genes_pseudobulk$gene_id, 
    markers = marker_genes_KRM_IDs
)

#pdf("../plots/SpatialDE_genes_analysis/upset_genes_combined_pooled.pdf", width = 6, height = 5)
upset(
    fromList(list_overlap), 
    order.by = "degree", 
    decreasing = FALSE, 
    sets = rev(names(list_overlap)), 
    set_size.show = TRUE, 
    set_size.scale_max = 4000, 
    keep.order = TRUE
)
#dev.off()


list_overlap <- list(
    SpatialDE_151673 = genes_SpatialDE_filt[[9]]$gene_id, 
    HVGs = genes_HVGs$gene_id, 
    pseudobulk_layers = genes_pseudobulk$gene_id, 
    markers = marker_genes_KRM_IDs
)

#pdf("../plots/SpatialDE_genes_analysis/upset_genes_combined.pdf", width = 6, height = 5)
upset(
    fromList(list_overlap), 
    order.by = "degree", 
    decreasing = FALSE, 
    sets = rev(names(list_overlap)), 
    set_size.show = TRUE, 
    set_size.scale_max = 2500, 
    keep.order = TRUE
)
#dev.off()
```


```{r, fig.width = 5.75, fig.height = 4.25}
# select gene(s)
genes_fig <- c("MOBP", "PCP4", "SNAP25", "HBB", "IGKC", "NPY")

# subset original SCE object
sce_fig <- sce[toupper(rowData(sce)$gene_name) %in% genes_fig, 
               colData(sce)$sample_name == "151673"]
dim(sce_fig)

# extract x-y coordinates of spots (note: y coordinate is reversed)
xy_coords <- data.frame(
    x_coord = colData(sce_fig)[, c("imagecol")], 
    y_coord = -colData(sce_fig)[, c("imagerow")]
)

# extract expression levels (UMI counts)
exprs_fig <- counts(sce_fig)
dim(exprs_fig)
# replace gene IDs with gene names for plots
stopifnot(all(rownames(exprs_fig) == rowData(sce_fig)$gene_id))
rownames(exprs_fig) <- rowData(sce_fig)$gene_name
# check column ordering matches spots
stopifnot(all(colData(sce_fig)$barcode == colnames(exprs_fig)))
stopifnot(length(colData(sce_fig)$barcode) == length(colnames(exprs_fig)))
# order genes alphabetically
exprs_fig <- exprs_fig[order(rownames(exprs_fig)), , drop = FALSE]

d_plot <- cbind(
    barcode = colData(sce_fig)$barcode, 
    xy_coords, 
    as.data.frame(as.matrix(t(exprs_fig)))
)

d_plot <- melt(
    d_plot, 
    id.vars = c("barcode", "x_coord", "y_coord"), 
    variable.name = "gene_name", 
    value.name = "UMIs"
)

d_plot$gene_name <- factor(d_plot$gene_name, levels = genes_fig)

d_plot$gene_type <- NA
d_plot$gene_type[d_plot$gene_name %in% c("MOBP", "PCP4", "SNAP25")] <- "laminar"
d_plot$gene_type[d_plot$gene_name %in% c("HBB", "IGKC", "NPY")] <- "non-laminar"
d_plot$gene_type <- factor(d_plot$gene_type, levels = c("laminar", "non-laminar"))

max_UMI <- max(d_plot$UMIs)

p_selected <- 
    ggplot(d_plot, aes(x = x_coord, y = y_coord, color = UMIs)) + 
        facet_wrap(~ gene_name) + 
        geom_point(size = 0.1) + 
        scale_color_gradientn(colors = c("gray95", "deepskyblue", "blue", "black"), 
                              values = rescale(c(0, 3, 15, max_UMI)), 
                              limits = c(0, max_UMI), 
                              name = "UMIs") + 
        coord_fixed() + 
        theme_bw() + 
            theme(strip.text.x = element_text(size = 13), 
                  legend.title = element_text(size = 13), 
                  legend.text = element_text(size = 13), 
                  axis.title.x = element_blank(), 
                  axis.title.y = element_blank(), 
                  axis.text.x = element_blank(), 
                  axis.text.y = element_blank(), 
                  axis.ticks.x = element_blank(), 
                  axis.ticks.y = element_blank(), 
                  panel.grid.major = element_blank(), 
                  panel.grid.minor = element_blank())

print(p_selected)

filename <- paste0("../plots/figure/Fig7_spatialDE_expression_selected.pdf")
ggsave(filename, width = 5.75, height = 4.25)
```


```{r, fig.width = 2.25, fig.height = 2.5}
# figure for schematic: showing PCP4 only
d_plot %>% 
    filter(gene_name == "PCP4") %>% 
        ggplot(., aes(x = x_coord, y = y_coord, color = UMIs)) + 
            geom_point(size = 0.2) + 
            scale_color_gradientn(colors = c("gray95", "deepskyblue", "blue", "black"), 
                                  values = rescale(c(0, 3, 15, max_UMI)), 
                                  limits = c(0, max_UMI), 
                                  name = "UMI counts") + 
            coord_fixed() + 
            theme_bw() + 
            theme(legend.position = "none", 
                  axis.title.x = element_blank(), 
                  axis.title.y = element_blank(), 
                  axis.text.x = element_blank(), 
                  axis.text.y = element_blank(), 
                  axis.ticks.x = element_blank(), 
                  axis.ticks.y = element_blank(), 
                  panel.grid.major = element_blank(), 
                  panel.grid.minor = element_blank())

filename <- paste0("../plots/figure/schematic/schematic_PCP4.pdf")
ggsave(filename, width = 2.25, height = 2.5)
```


## Session information

```{r}
sessionInfo()
```



